---
title: "IP Weighting for Categorical Exposures"
author: "Ashley I Naimi & Brian W Whitcomb"
date: "`r paste0('Spring ', 2024)`" #format(Sys.Date(), '%Y')
urlcolor: blue
link-citations: yes
output: html_document
---

# Importing Data

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=40),tidy=TRUE)

```

This file serves as the supplementary material to *Inverse Probability Weighting for Categorical Exposures* published in the *American Journal of Epidemiology* in the AJE Classroom Section. The purpose is to demonstrate how to construct inverse probability weights for a categorical exposure using the R programming language.

We begin with importing the NHEFS data:

```{r tidy = F, warning = F, message = F}

#' Install and load relevant packages (using pacman)
pacman::p_load(
  rio,          
  here,         
  skimr,        
  tidyverse,     
  lmtest,
  sandwich,
  broom,
  ggpubr
  )

thm <- theme_classic() +
  theme(
    legend.position = "top",
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )
theme_set(thm)

#' Data are available at the following URL
file_loc <- url("https://bit.ly/47ECRcs")

#' This begins the process of cleaning and formatting the data
nhefs <- read_csv(file_loc) %>%
  select(wt82_71, exercise,
         sex, age, race, income, school) %>%
  mutate(income=as.numeric(income>15)) %>%
  na.omit(.) # use complete case to simplify

nhefs

```

# IP Weighting for Categorical and Continuous Exposures

To construct weights for a categorical exposure, we can use multinomial logistic regression to obtain the predicted probability of being in the **observed exposure category.** Note that many other modeling approaches can be used, depending on the estimand of interest and the assumptions one is willing to make about the propensity score model. However, multinomial logistic regression is a relatively flexible approach that will yield propensity score estimates usable in most settings. 

In R, multinomial regression can be deployed using the `VGAM` package via the `vglm` function. If we assume that sex, age, race, income, and education (school) are confounders of the exercise and weight change relation, we can fit the following propensity score model:


```{r tidy = F, warning = F, message = F}

pacman::p_load(VGAM)

ps_mod <- vglm(factor(exercise) ~ sex + age + race + income + school, 
               data = nhefs, 
               family = "multinomial")

summary(ps_mod)

```

Once this propensity score model is fit, we need to obtain predicted probabilities of being exposed in each category:

```{r tidy = F, warning = F, message = F}

ps_matrix <- predict(ps_mod, type = "response")

ps_matrix <- data.frame(ps_matrix)

names(ps_matrix) <- c("pEx0","pEx1","pEx2")

nhefs <- cbind(ps_matrix, nhefs)

ps_matrix <- data.frame(ps_matrix, exposure = nhefs$exercise)

```

Let's look at the summary distributions of each conditional probability:

```{r}

summary(ps_matrix[,1])

summary(ps_matrix[,2])

summary(ps_matrix[,3])

```

We can also explore the overlap in these propensity scores:

```{r tidy = F, warning = F, message = F}

plot_dat1 <- data.frame(ps_matrix[,c(1,4)], pEx = "pEx0")
plot_dat2 <- data.frame(ps_matrix[,c(2,4)], pEx = "pEx1")
plot_dat3 <- data.frame(ps_matrix[,c(3,4)], pEx = "pEx2")

names(plot_dat1) <- names(plot_dat2) <- names(plot_dat3) <- c("propensity_score", 
                                                              "Exposure", 
                                                              "pEx")

plot_dat <- rbind(plot_dat1,
                  plot_dat2,
                  plot_dat3)

dim(plot_dat)
dim(ps_matrix)

head(ps_matrix, 10)

head(plot_dat, 10)

plot_dat0 <- plot_dat %>% 
  filter(pEx == "pEx0") %>% 
  mutate(Exposure = factor(Exposure))

p1 <- ggplot(plot_dat0) +
  geom_density(aes(x = propensity_score,
                   group = Exposure,
                   fill  = Exposure),
               bw = .03, alpha = .25) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = c("black", "gray", "white")) +
  xlab("Propensity Score: Pr(Ex = 0 | C)") + ylab("Density")

plot_dat1 <- plot_dat %>% 
  filter(pEx == "pEx1") %>% 
  mutate(Exposure = factor(Exposure))

p2 <- ggplot(plot_dat1) +
  geom_density(aes(x = propensity_score,
                   group = Exposure,
                   fill  = Exposure),
               bw = .03, alpha = .25) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = c("black", "gray", "white")) +
  xlab("Propensity Score: Pr(Ex = 1 | C)") + ylab("Density")

plot_dat2 <- plot_dat %>% 
  filter(pEx == "pEx2") %>% 
  mutate(Exposure = factor(Exposure))

p3 <- ggplot(plot_dat2) +
  geom_density(aes(x = propensity_score,
                   group = Exposure,
                   fill  = Exposure),
               bw = .03, alpha = .25) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1)) +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_manual(values = c("black", "gray", "white")) +
  xlab("Propensity Score: Pr(Ex = 2 | C)") + ylab("Density")

ggarrange(p1, p2, p3, ncol = 1, labels = c("a)", "b)", "c)"))

ggsave(here("figures", "ps_overlap_plot_v2.png"),
       width = 8, height = 20, units = "cm")

```

This is a propensity score overlap plot for the categorical (three level) exposure, and it does not suggest problematic overlap. For a three category exposure, we can also use ternary plots to evaluate **joint** overlap in the distribution of the propensity score. Details on ternary plots are provided in the eAppendix of Yoshida et al (2019) Pharmacoepidemiol Drug Saf, Jul; 28(7): 934–941, section 3.1.

In R, ternary plots can be obtained using the `ggtern` package, which is available on CRAN. Documentation and examples are available [here](http://www.ggtern.com/d/2.2.0/demo-points.html). 

We can use the `ps_matrix` data object to create our ternary plot. The basic formulation is as follows:

```{r tidy = F, warning = F, message = F}

pacman::p_load(ggtern)

# ternary plot
ggtern(data=ps_matrix,
       aes(x=pEx0,y=pEx1, z=pEx2)) +
  geom_point(aes(group = factor(exposure), 
                 color = factor(exposure)))

```

This plot shows how the probabilities of being in each exposure category are jointly related among those in each observed exposure category. Further details, including how to convert propensity scores to "preference scores" (which account for the baseline prevalence of each exposure category) are available in the paper by Yoshida et al. 


# Computing the Effect Contrasts

To construct stabilized weights used to estimate each of the parameters of interest outlined in the Classroom article, we need to obtain the predicted probability of the observed exposure from the propensity score matrix object:

```{r tidy = F, message = F, warning = F}

head(ps_matrix)

pscore_obs <- NULL
for(i in 1:nrow(ps_matrix)){
  pscore_obs <- rbind(pscore_obs,
                  ps_matrix[i, ps_matrix[i,]$exposure + 1] # note the "+ 1"
                  )
}

head(pscore_obs)

```

Each row in the `pscore_obs` object represents the predicted probability for each person being in their **observed exposure level**, conditional on the covariates in the propensity score model. With this, we can construct inverse probability weight as we did above. Note that we can stabilize these weights by including the overall mean of the observed exposure category. We can actually do this simply with an intercept only multinomial model, as we did above:

```{r tidy = F, warning = F, message = F}

mean(nhefs$exercise == 0)
mean(nhefs$exercise == 1)
mean(nhefs$exercise == 2)

ps_model <- vglm(factor(exercise) ~ 1,
               data = nhefs,
               family = "multinomial")

ps_num <- cbind(predict(ps_model, type = "response"),  nhefs$exercise)

ps_num <- data.frame(ps_num)

names(ps_num) <- c("p0", "p1", "p2", "exercise_level")

head(ps_num)

pscore_num <- NULL
for(i in 1:nrow(ps_num)){
  pscore_num <- rbind(pscore_num,
                      ps_num[i, ps_num[i,]$exercise_level + 1]
  )
}

head(pscore_num)

head(pscore_obs)

nhefs$sw_exercise <- pscore_num/pscore_obs

head(nhefs)

summary(nhefs$sw_exercise)

```

With these weights we can now estimate the average treatment affect between exercise and our outcome of interest (in this case, weight change). We use the `coeftest` function from the `lmtest` package along with the `sandwich` function to obtain robust variance estimates. 

```{r tidy = F, warning = F, message = F}

modelRD_ex1 <- lm(wt82_71 ~ relevel(factor(exercise),
                                    ref = "2"),
                 weight = sw_exercise,
                 data = nhefs)

modelRD_ex2 <- lm(wt82_71 ~ relevel(factor(exercise),
                                    ref = "1"),
                 weight = sw_exercise,
                 data = nhefs)

coeftest(modelRD_ex1, vcov. = vcovHC(modelRD_ex1, "HC3"))
coefci(modelRD_ex1, vcov. = vcovHC(modelRD_ex1, "HC3"))

coeftest(modelRD_ex2, vcov. = vcovHC(modelRD_ex2, "HC3"))
coefci(modelRD_ex2, vcov. = vcovHC(modelRD_ex2, "HC3"))

```

We also deploy the bootstrap here to account for variability in the propensity score model:

```{r tidy = F, warning = F, message = F}

pacman::p_load(boot)

boot_func <- function(data, index){
  
  boot_dat <- data[index,]
  
  # numerator model
  num_mod <- vglm(factor(exercise) ~ 1, 
                 data = boot_dat, 
                 family = "multinomial")
  
  # denominator model
  ps_mod <- vglm(factor(exercise) ~ sex + age + race + income + school, 
                 data = boot_dat, 
                 family = "multinomial")
  
  # denominator
  ps_matrix <- data.frame(predict(ps_mod, type = "response"), 
                          exposure = boot_dat$exercise)
  pscore_obs <- NULL
  for(i in 1:nrow(ps_matrix)){
    pscore_obs <- rbind(pscore_obs,
                    ps_matrix[i, ps_matrix[i,]$exposure + 1] # note the "+ 1"
                    )
  }
  
  # numerator
  ps_matrix <- data.frame(predict(num_mod, type = "response"), 
                          exposure = boot_dat$exercise)
  
  pscore_num <- NULL
  for(i in 1:nrow(ps_matrix)){
    pscore_num <- rbind(pscore_num,
                        ps_matrix[i, ps_matrix[i,]$exposure + 1]
    )
  }
  
  boot_dat$sw_exercise <- pscore_num/pscore_obs
  
  modelRD_ex1 <- lm(wt82_71 ~ relevel(factor(exercise),
                                      ref = "2"),
                   weight = sw_exercise,
                   data = boot_dat)
  
  modelRD_ex2 <- lm(wt82_71 ~ relevel(factor(exercise),
                                      ref = "1"),
                   weight = sw_exercise,
                   data = boot_dat)
  
  res1 <- summary(modelRD_ex1)$coefficients[2:3, 1]
  res2 <- summary(modelRD_ex2)$coefficients[2, 1]
  
  return(c(res1, res2))
}

boot_res <- boot::boot(nhefs, boot_func, R = 2000)

boot.ci(boot_res, type = "norm", index = 1)

boot.ci(boot_res, type = "norm", index = 2)

boot.ci(boot_res, type = "norm", index = 3)

```


# References